---
title: "Project Title Here"
author: "Group Member Names"
format: pdf
editor: visual
---

```{r,include=FALSE}
library(tidyverse)
library(openintro)
library(Stat2Data)
library(infer)

```

## Abstract

Delete this text and write the abstract for your paper here.

## Introduction

Please delete this text and write your introduction here. Please follow the instructions on the project description on the \[course website\]().

### Research Questions & Hypotheses

We sought to answer the following research questions:

1.  

2.  

These are the research hypotheses:

1.  

2.  

## Literature Review

Write your literature review here including in-text citations. See the project description on the course website for further details.

## Methodology

### Sample and Data Collection Methods

Describe your sample and sampling methods used. Provide details.

Describe how the data may have been collected (e.g., surveys, etc). See project description for more details.

### Description of Data Analysis Techniques

Describe how you analysed data for your study. Please see project description on the course website for details of what goes here.

## Data Analysis

### Exploratory Analyses

Run your exploratory analyses here. Include interpretations of software output.

```{r, echo=FALSE}

```

```{r, echo=FALSE}

```

```{r, echo=FALSE}

```

### Inferential Analyses

Run your inferential analyses here for each research question. Include interpretations of software output. Add more code chunks if necessary.

```{r, echo=FALSE}



```

```{r, echo=FALSE}



```

```{r, echo=FALSE}


```

## Conclusion

Write the conclusion here. Be sure to look at the project description.

## References

Have a list of sources cited in your work. This could be journal articles, books, websites, conference proceedings, etc.
