---
categories: "Lab 4"
title: "Lab 4: Introduction to Linear Regression"
execute:
  eval: true
format:
  html: 
    code-link: true #Supposed to create a link to documentation but its not, for some reason.
    code-block-border-left: "#31BAE9"
    code-fold: show #also try true OR false and decide if you like it.
    code-tools: false
    code-copy: true
---

```{r, include=FALSE, echo=F}
library(tidyverse)
library(openintro)
library(broom)
```

```{r, include=F, echo=F}
data(duke_forest)
data("cars04")
```

## [Introduction]{style="color:#666633;"}

In this lab, we will explore the concept of simple linear regression, a fundamental statistical technique used to model the relationship between a predictor (aka independent) and a response (dependent) variable. We call this **simple linear regression** because it involves only one predictor variable and one response variable. The term linear suggests that we will consider relationships that show a linear pattern. Statistical techniques for modelling non-linear relationships and/or that use multiple variables (predictor and even response) do exist but are beyond the scope of this course. Linear regression is widely used in various fields, including economics, biology, engineering, and social sciences, to make predictions and understand relationships between variables.

## [Creating a quarto file]{style="color:#666633;"}

Go to posit cloud (<https://posit.cloud>) and open the MATH 144 project. Once your MATH 144 project is open, clean up the console and the environment area. You may also want to close all open files.

Next, create your new quarto file as follows:

- File \> New File \> Quarto document. Use the file name `Lab_4`. Do not delete the `.qmd` extension.

- Add the `author` and `title` keys to the `YAML` section. Write your name (in quotes) in the author field (`author: "Student Name"`) and set the title to "`Simple Linear Regression"`.

- Change the output option to `pdf` then click on `render`. A PDF document with your name and document title should pop up.

## [Packages]{style="color:#666633;"}

You will need `openintro`, `tidyverse`, and `broom` packages for this lab. We have not installed the `broom` package before, so you will need to do that first. To do this, you can run the code `install.packages("broom")` in **the console**. After installing the `broom` package, create a level 2 header and call it Packages. Under this, insert a new code chunk and load the three packages using the `library()` function as follows:

``` toml
library(openintro)
library(tidyverse)
library(broom)
```

Be sure to run this chunk to make sure the packages were successfully loaded. For a cleaner looking document, use `include = FALSE` in the code chunk to prevent the output of this chunk.

You may use the following alternative way of deactivating the chunk output. If you use this, delete the first package code chunk.

``` toml
#| include: false
library(openintro)
library(tidyverse)
library(broom)
```

## [Loading the Data]{style="color:#666633;"}

We will continue using the `duke_forest` data set contained in the `openintro` package. Insert a new code chunk and load the data using the code below:

```{r}
data(duke_forest)
```

This data were used in previous labs, but you can refresh your memory by running the query `?duke_forest`. It is **very important** that you examine this data set first if you don't remember it well.

## [Exploring Relationships]{style="color:#666633;"}

We are interested in the variables `price` (given in US dollars) and `area` of the home (in square feet). Specifically, we want to explore the relationship between the two variables and create linear regression model with `area` as the predictor variable and `price` as the response variable.

Below is a scatter plot showing the relationship between `price` and `area`:

```{r}
ggplot(data = duke_forest) +
  aes(x = area, y = price) +
  geom_point()
```

**Question:** Describe the relationship between `price` and `area` based on the scatter plot. Be sure to talk about the direction, strength, and form of the relationship. Are there any unusual observations? What might be a reasonable explanation for these unusual observations?

**Model Answer:** The relationship between `price` and `area` appears to be positive — large homes are generally more expensive and vice versa. The overall relationship appears to be linear and moderately strong, suggesting that a linear regression model may be appropriate for this data. There seems to be a few unusual observations (lower right), particularly homes with very low prices relative to their area. These outliers could be due to factors such as location, amenities, or unique features of the homes that are not captured by the scatter plot.

## [Correlation coefficient]{style="color:#666633;"}

To quantify the strength and direction of the linear relationship between `price` and `area`, we use a statistic known as the correlation coefficient (abbreviated as $r$\$r\$. We calculate this in R using the `cor()` function as shown below:

```{r}
duke_forest |>
  summarize(cor(price, area))
```

**Note:** The correlation value of 0.67 confirms our earlier assessment of the relationship from the scatter plot.

## [Visualizing the Model]{style="color:#666633;"}

If you want to see the **least squares regression line**, you simply add the `geom_smooth()` function to the scatter plot created earlier. You will also use the argument `method` and set it to `"lm"` as shown below:

```{r}
ggplot(data=duke_forest, aes(x=area, y=price))+
  geom_point()+
  geom_smooth(method="lm") # this adds the line
```

## 

## [Finding the Regression Equation]{style="color:#666633;"}

To find the equation of the line, we can use the `lm()` function as shown below. The first argument takes the form `y ~ x,` where `y` is the response variable and `x` is the predictor variable.

```{r}
lm(price~area, data=duke_forest)

```

It is helpful to save the model as an object so that when you need it, you don't write new code. The code below saves the model as `model` and also prints the model right under:

```{r}
model <- lm(price~area, data=duke_forest) # save model
print(model) # to show model
```

We can write the equation of the model for predicting price as follows:

$$\widehat{price} = 159.5\times (area) + 116,652.3$$

**Interpretation of slope and intercept**

- ***Intercept:*** The intercept is 116,652.3 and its units are dollars. It means that on average, homes with 0 square feet area would cost \$116,652.3. This does not have any practical interpretation in this context since you cannot have a home measuring 0 square feet.
- ***Slope:*** The slope is 159.5 and its units are ***dollars per square feet***. The slope means that on average, an additional square foot in home area results to a \$159.5 increase in sale price.

## [Model Evaluation]{style="color:#666633;"}

We can use our model to make predictions. For example, if we wanted to predict the sale price of a 2,000 square feet home, we can use the equation as follows:

$$\widehat{price} = 116,652.3+159.5\times 2000 = \$436,652.3$$

Our model predicts that a home with 2,000 square feet in the same area where the data were collected would sell for approximately \$436,652.3.

The `augment()` function can be used to calculate the predictions (estimated home price) for each home in the data for which the area is available. The head function is used to view the first few rows of the new object.

```{r}
predictions <- augment(model)
head(predictions)
```

We are concerned with `.fitted` and `.resid` columns only. The `.fitted` column show the prices calculated based on the model (the predictions) while the `.resid` column shows the difference between `actual` and `predicted` prices. Residuals are used in model evaluation. A small residual indicates that the model's prediction is close to the actual value, while a large residual indicates a poor prediction.

To visualize the residuals, we can create a residual plot. A residual plot displays the residuals on the y-axis and the predictor variable (in this case, `area`) on the x-axis. Ideally, we want to see a random scatter of points around the horizontal line at 0, which indicates that our model is appropriate for the data. If we see patterns in the residuals, it may suggest that our model is not capturing all the information in the data.

```{r}
ggplot(data=predictions, aes(x=area, y=.resid))+
  geom_point()+
  geom_hline(yintercept=0, linetype="dashed", color="red")
```

We can also create a histogram of the residuals to check if they are approximately normally distributed. This is another assumption of linear regression.

```{r}
ggplot(data=predictions, aes(x=.resid))+
  geom_histogram()+
  labs(title="Histogram of Residuals", x="Residuals", y="Frequency")
```

A histogram that is roughly bell-shaped suggests that the residuals are approximately normally distributed, which is a good sign for our linear regression model. If the histogram shows a skewed distribution or has multiple peaks, it may indicate that our model is not appropriate for the data.

In summary, certain assumptions/conditions need to be checked when performing linear regression. These include:

- ***Linearity:*** The relationship between the predictor and response variable should be roughly linear. This can be checked using the scatter plot.
- ***Independence:*** The residuals should be independent of each other. This is generally assumed if the data were collected randomly.
- ***Homoscedasticity:*** The residuals should have constant variance across all levels of the predictor variable. This can be checked using the residual plot. The residual plot above does not show serious violations of this assumption.
- ***Normality:*** The residuals should be approximately normally distributed. This can be checked using the histogram of residuals.

## [In the Literature]{style="color:#666633;"}

Have a section here illustrating how analyses are reported in the literature. Can use the APA format because that is what the projects require.

## [Summary]{style="color:#666633;"}

In this lab, we explored the concept of simple linear regression using the `duke_forest` data set. We created a scatter plot to visualize the relationship between `price` and `area`, calculated the correlation coefficient to quantify the strength and direction of the relationship, fitted a linear regression model, visualized the model, and evaluated its performance using residuals. Linear regression is a powerful tool for understanding relationships between variables and making predictions based on those relationships.

## [Exercises]{style="color:#666633;"}

Create a new header called **Lab 4 Exercises** and answer the questions below. When you are done, render the document to generate a PDF. Download the document and submit it on Canvas under Lab 4. The lab will be graded out of 20 (Ten points for exploration work above, and 10 for the exercises).

### [Question One]{style="color:#666633;"}

**(1 pt)** The data set `cars04` is in the openintro package. Load the data set into your work space. What are the cases in this data set and how many are they? What do the variables `weight` and `eng_size` represent, and what are the units for these variables?

### [Question Two]{style="color:#666633;"}

**(2 pts)** Create a scatter plot to visualize the relationship between `eng_size` (x-axis) and `weight` (y-axis). Describe the relationship between the two variables, making sure to talk about the direction, strength, and form of the relationship. State if there are any unusual observations and provide a reasonable explanation for the unusual observations.

### [Question Three]{style="color:#666633;"}

**(1 pt)** Calculate the correlation coefficient between `eng_size` and `weight`. Comment on this statistic.

### [Question Four]{style="color:#666633;"}

**(3 pts)** Generate a linear regression model for predicting `weight` based on `eng_size` (predictor). Write the equation of the model and interpret the `slope` and `intercept` in the context of the data. Be sure to state the units of the `slope` and `intercept`.

### [Question Five]{style="color:#666633;"}

**(1 pt)** Use your model from #4 above to estimate the weight of a car with an engine size of 3.4 liters. Explain/show work.

### [Question Six]{style="color:#666633;"}

**(2 pts)** What's the residual for the first case in the data set? For this case, does the model overestimate or underestimate the weight of the car? Explain how you know this.
