---
categories: "Lab 6"
title: "Lab 6: Introduction to Inference - Hypothesis Testing"
execute:
  eval: true
format:
  html: 
    code-link: true #Supposed to create a link to documentation but its not, for some reason.
    code-block-border-left: "#31BAE9"
    code-fold: show #also try true OR false and decide if you like it.
    code-tools: false
    code-copy: true
---

```{r, include=FALSE, echo=F}
library(tidyverse)
library(openintro)
library(infer)
data(sex_discrimination)
data(opportunity_cost)

```

```{r, include=F, echo=F}
data(duke_forest)
```

## [Introduction]{style="color:#4166f5;"}

In this lab, we will introduce statistical inference using a popular package known as `infer` (short form for inference). Specifically, we will conduct a hypothesis test for difference in two proportions via randomization. In future labs, we will explore other types of inference such as confidence intervals.

## [Creating a quarto file]{style="color:#4166f5;"}

Go to posit cloud (<https://posit.cloud>) and open the MATH 144 project. Once your MATH 144 project is open, clean up the console and the environment. Close any open documents and datasets that you don't need.

Next, create your new quarto file as follows:

- File \> New File \> Quarto document. Use the file name `Lab_6`. Do not delete the `.qmd` extension.

- Add the `author` and `title` keys to the `YAML` section. Write your name (in quotes) in the author field and set the title to "`Introduction to Statistical Inference - Hypothesis Testing"`.

- Change the output option to `pdf` then click on `render`. A PDF document with your name and title should pop up. If the popup blocker shows up, click try again.

## [Research Question & Scenario]{style="color:#4166f5;"}

***Here is the Context/Scenario:***

Forty eight male bank supervisors were each given identical resumes to decide whether the candidate should be promoted or not. Twenty four of the files were labelled as male, meaning the applicant was male, while the other 24 were labelled as female.

We seek to answer the following research question via hypothesis testing:

***Research Question:*** "Are individuals who identify as female discriminated against in promotion decisions made by their managers who identify as male?”

## [Package and Data]{style="color:#4166f5;"}

The data set is called `sex_discrimination` and is available in the `openintro` package. Apart from `openintro`, we will need the `tidyverse` and `infer` packages. The `infer` package contains a wide range of functions including functions for running simulations and calculating statistics. Since we have installed `openintro` and `tidyverse` before, we only need to install `infer`. Using **the console**, run the code below to install the `infer` package:

``` toml
install.packages("infer")
```

Now, create a new code chunk and use the code below to load packages and the data (sex_discrimination).

``` toml
#| include: false
library(tidyverse)
library(openintro)
library(infer)
data(sex_discrimination)
```

Run the above code. If everything runs correctly, a new object (sex_discrimination) will pop up in the environment area. This is the data set. Open the data set and examine it before you proceed. Check what the cases are and how many there are. Check the variables and determine the predictor and response variables making sure this aligns with the research question.

## [Hypotheses Statements]{style="color:#4166f5;"}

We write the null and alternative hypothesis in words and in symbols as follows:

***Null Hypotheis in words:*** There is no difference in promotion rates (proportions) between male and female candidates *in the population* (i.e., There is no discrimination against female candidates in promotion decisions and the observed difference in our sample is due to random chance).

In symbols, we write the null hypothesis as

$$H_0 : p_M-p_F=0$$

where, $p_M$= proportion of male candidates that would be promoted in the population, and $p_F$= proportion of female candidates that would be promoted.

***Alternative Hypothesis in words:*** There is a difference in promotion rates between male and female candidates in favor for males (i.e., There is discrimination against female candidates in the population).

In symbols, we write the null hypothesis as

$$H_A : p_M-p_F>0$$

We will run a hypothesis test to see if the sample data provides strong enough evidence to reject the null hypothesis.

## [Observed Statistic]{style="color:#4166f5;"}

When performing statistical inference, it is very important to be clear on the statistic of interest. In the case of the sex discrimination study, the statistic of interest is ***the difference in proportions*** between male and female candidates. We use the letter $\hat{p}$ (p-hat) to denote the sample proportion (a statistic) and $p$ to denote the population proportion (a parameter). Notice that the hypotheses statements use population parameters (in this case $p$). In the hypotheses above, $p_M$ is the promotion rate for males male candidates in the population while $p_F$ is the promotion rate for female candidates.

The ***observed statistic*** in this case is the difference in promotion rates in ***the sample***. We use this observed difference as an estimate of the true population difference (note that we do not know what the true difference is).

$$Obs\_Stat=\hat{p}_M-\hat{p}_F$$

In R, we can use the `count()`, and `group_by()` functions to create a summary table as shown below:

```{r}
sex_discrimination |>       #get your data
  count(sex, decision) |>   #count cases in both variables
  group_by(sex) |>          #group cases by sex (predictor)
  mutate(prop = n / sum(n)) #calculate the proportions
```

From the table, we can easily compute the observed statistic as follows:

\begin{align}
\hat{p}_{diff}&=\hat{p_1}-\hat{p_2}\\
&=0.875-0.5833\\
&= 0.2917
\end{align}

Since these are two categorical variables, we can visualize the distribution of the data using a side-by-side bar plot as shown below:

```{r}
sex_discrimination |>
  ggplot(aes(x = sex, fill = decision)) +
  geom_bar(position = "dodge")
```

## [Running Simulations]{style="color:#4166f5;"}

There are 4 main functions in the `infer package` that we will use over and over in coming days and weeks. The functions are:

- `specify()`***:*** this allows you to specify the variables and the relationship you are interested in. In this case, `sex` is the predictor and `decision` is the response.

- `hypothesize()`***:*** This is where you specify the nature of the null hypothesis.

- `generate()`***:*** runs the simulations that you are interested in. We will set this to `permute` which is a way of randomizing the data.

- `calculate()`***:*** allows you to calculate statistic of interest for each simulation. In this case `diff in props` (difference in proportions).

Let us now use the infer package to run the simulations for the sex discrimination case study. We save the simulated data as ***simulated_diffs***. We will need this data in later stages.

```{r}
simulated_diffs <- sex_discrimination |>
		      specify(decision ~ sex, success = "promoted") |>
			    hypothesize(null = "independence") |>
			    generate(reps = 1000, type = "permute") |>
			    calculate(stat = "diff in props", order = c("male", "female"))
```

> Under specify(), the order is response\~predictor, and the success level is set to "promoted".
>
> Under hypothesize(), we set null to "independence" because our null hypothesis assumes the two variables are not associated.
>
> Under generate(), we want to do 1000 repeats, and the type of shuffling is called "permute".
>
> Under calculate(), we state the statistic we want and the order in which we want the difference done.

After running the above code, a new object will pop up in the environment area. This object (aka null distribution) contains the 1000 differences calculate via simulation. Click on the object or use the `head()` function to view the first few rows of the data set.

```{r}
head(simulated_diffs)
```

## [Visualizing the Null Distribution]{style="color:#4166f5;"}

Next, we want to create a histogram to visualize the distribution of the differences computed above. Remember we stored the differences as `simulated_diffs`. We will also place a vertical line on the histogram to see the location of the observed statistic (the original difference of 0.2917) relative to the graph. We use the code below:

```{r, message=F}
			ggplot(data=simulated_diffs, aes(x = stat)) + 
			geom_histogram()+
			geom_vline(aes(xintercept = .2917), color = "red")
```

Note that this histogram is called the ***null distribution*** because it represents the distribution of differences we would expect to see if the null hypothesis were true. The null distribution is centered around 0 because under the null hypothesis, we expect no difference in promotion rates between the sexes.

## [Getting the P-Value]{style="color:#4166f5;"}

The P-value is the percent of values that are at least as extreme as the observed statistic (i.e., the red line). It is the probability of observing a difference as extreme as 0.2917 or more extreme assuming that the null hypothesis is true (i.e., that there is no discrimination going on). The infer package comes with a function known as `get-p-value()` for computing the p-value from the ***null distribution*** (i.e., our simulated_diffs). See code below:

```{r}
simulated_diffs |>
			get_p_value(obs_stat = 0.292, direction = "greater")
```

Here, the argument `obs_stat` is used to set the observed statistic, while `direction` is for specifying the alternative hypothesis. Since our observed statistic is to the right of the center, and our alternative hypothesis is directional, we set the direction argument to "greater". If the alternative hypothesis was bidirectional or two-sided (i.e., $p_M-p_F \neq 0$), we would set direction to `two-sided`, which means we would be looking for values that are as extreme as 0.2917 in either direction (both left and right tails of the distribution). This would essentially double the P-value we get from the right tail.

## [Conclusion in Context]{style="color:#4166f5;"}

The P-value obtained is 0.003 (or 0.3%). This means that if there was no discrimination going on (i.e., null hypothesis were true), we would would expect to see a difference of at least 0.2917 about 0.3% of the time. This low probability indicates that observing such a large difference from chance alone is very rare. This means that the data provides strong evidence against the null hypothesis and we can conclude that discrimination is likely going on against the female candidates.

It is common to use a P-value cut off point of 0.05 (5%) to decide whether the data provides strong evidence against the null hypothesis. This cut off point is called the ***discernibility level***. In general, when the P-value is less than the ***discernibility level (0.05)*** we reject the null hypothesis in favor of the alternative. When the P-value is greater than the discernibility level, we fail to reject the null hypothesis and can't make a decision either way.

## [Exercises]{style="color:#4166f5;"}

Create a new level 1 header called **Lab 6 Exercises** and answer the questions below. Each question should be labelled appropriately (e.g., **Question 1**) using level 2 headers. No need to retype or copy and paste the questions. When you are done, render the document to generate a PDF. Download the document and submit it on Canvas under Lab 6. The lab will be graded out of 20 (Ten points for exploration work above, and 10 for the exercises). The lab exploration work must have headers and descriptive text describing what each code chunk is achieving. Avoid copying and pasting without reading.

***Scenario:***

One-hundred and fifty students were recruited in a study that aimed to understand whether reminding college students about saving money for later would make them more thriftier. The students were split into 2 equal groups (control and experiment). Each student was given the following prompt:

*"Imagine that you have been saving some extra money on the side to make some purchases, and on your most recent visit to the video store you come across a special sale on a new video. This video is one with your favorite actor or actress, and your favorite type of movie (such as a comedy, drama, thriller, etc.). This particular video that you are considering is one you have been thinking about buying for a long time. It is available for a special sale price of \$14.99. What would you do in this situation? Buy video or Not buy video?"*

Students in the control group were given the following two options:

A.  Buy video
B.  Not buy video

Students in the experiment group were given a slightly modified option B:

A.  Buy video
B.  Not buy this video. *Keep the \$14.99 for other purchases*.

**Research Question:** Does reminding students about saving money for later make them more thriftier (i.e., less likely to buy the video)?

The data set is called `opportunity_cost` and is available in the `openintro` package.

1.  **(1 pts)** Load the `opportunity_cost` data set into your workspace. Identify the predictor and response variables and classify each as categorical or numerical. Be sure to write down the variable names exactly as they are saved are in the data set.
2.  **(1 pts)** State the appropriate null and alternative hypotheses for answering the research question.
3.  **(2 pts)** Create a summary table that shows the counts and proportions of students and their buying decisions in each group. Use the summary table to calculate the ***observed statistic***.
4.  **(2 pts)** Use the `infer` package functions to run 1000 simulations under the null hypothesis and save the results as **simulated_diffs**. This is your ***null distribution***. Be sure to specify the correct variables and relationship under `specify()`, the correct nature of the null hypothesis under `hypothesize()`, the correct type of shuffling under `generate()`, and the correct statistic under `calculate()`.
5.  **(2 pts)** Create a histogram to visualize the ***null distribution*** created in #4 above. Be sure to place a vertical line to show the location of the ***observed statistic*** from #3. Why is the null distribution centered where it is?
6.  **(2 pts)** Use the `get-p-value()` function to find the P-value. Is the p-value large or small, and what does that mean in context of the data? Note that this is where you make a decision about the null hypothesis.
